---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/vineclust)](https://CRAN.R-project.org/package=vineclust)
[![R-CMD-check](https://github.com/oezgesahin/vineclust/workflows/R-CMD-check/badge.svg)](https://github.com/oezgesahin/vineclust/actions)
[![codecov](https://codecov.io/gh/oezgesahin/vineclust/branch/main/graph/badge.svg?token=EUTCCS3B5U)](https://codecov.io/gh/oezgesahin/vineclust)
<!-- badges: end -->

# vineclust: Model-Based Clustering with Vine Copulas

An R package that maps model-based clustering across continuous non-Gaussian multivariate data using multidimensional Vine Copulas. Originally proposed in the [VCMM Algorithm](https://arxiv.org/pdf/2102.03257.pdf), `vineclust` excels in detecting data sub-populations spanning significant skewness and asymmetric dependencies.

> **v0.2+ Update**: The core engine of `vineclust` has been migrated to `rvinecopulib`! This guarantees significantly faster copula estimations, entirely vectorized log-likelihood functions, tracking hooks via `progressr`, and extensive capability for non-standard copula models like the **nonparametric** bivariate distributions.

It depends on [rvinecopulib](https://github.com/rvinecopulib/rvinecopulib), [fGarch](https://github.com/cran/fGarch), [mclust](https://github.com/cran/mclust), and [univariateML](https://github.com/JonasMoss/univariateML).

## Installation

You can install the development version from [GitHub](https://github.com/oezgesahin) with:

```r
# install.packages("remotes")
remotes::install_github("oezgesahin/vineclust")
```

---

## đź“– Wiki: Why & How to Use Vine Copula Clustering

Standard model-based clustering algorithms (such as Gaussian Mixture Models - GMMs) operate under the assumption of multivariate normality. This means they intrinsically impose linear constraints. For naturally skewed sets like financial logs, genomic matrices, or environmental analytics, this approach routinely misclassifies entities or artificially splinters actual clusters to compensate for asymmetric dependencies.

**Vine Copulas correctly resolve this by adhering to Sklar's Theorem:**
1. They independently target and fit the univariate behavior (heavy tails, skewed normals, logic peaks) separating it precisely from the multi-variable connection graph.
2. They map out the cross dependencies securely into hierarchical bivariate building blocks (Vines).

### How it operates mathematically
`vineclust` relies upon the Expectation-Conditional Maximization (ECM) framework.
1. **Initial Assignments**: Extracts preliminary cluster centroids relying on heuristic paths like `K-Means` or standard `mclust` Gaussian mixtures.
2. **E-Step**: Calculates expectation probabilities (`z_values`) for observational assignments based on current densities. 
3. **CM-Steps (Conditional Maximizing)**: Numerically maps likelihoods fitting highly specific margin rules (Cauchy, Lognormal) and re-evaluating the central `rvinecopulib` edge dependencies accurately. 

---

## Package Overview & Workflow

* `vcmm()`: The engine's core model fitting action. Fits the entire vine copula network targeting your `data` for an N `total_comp`. Generates an object of class `vcmm_res()`.
  * `print()`: Exposes high-level model mechanics (BIC, LogLik, Total Pars).
  * `summary()`: Unpacks deep array metrics involving univariate properties, edge classes, and mixture proportions.
* `dvcmm()`, `rvcmm()`: Secondary tools used for retrieving internal density probabilities or launching random synthetic arrays tied to built outputs.

### Marginals Support
We bridge distinct functional patterns for unimodal properties:
`cauchy`, `gamma`, `llogis`, `lnorm`, `logis`, `norm`, `snorm`, `std`, `sstd` (skew Student's t).

### Copula Network Families
`vineclust` adopts the universal string architecture inherent in `rvinecopulib`. By declaring properties such as `"all"`, `"parametric"`, or `"nonparametric"`, models instantly cycle through configurations to test properties like Rotated Archimedean (Clayton, Gumbel, Frank, Joe, BB1/6/7/8) without legacy rigid integer constraints.

---

## Usage

```{r data_load, eval=FALSE}
library(vineclust)
library(progressr)

# data from UCI Machine Learning Repository 
data_wisc <- read.csv("http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/wdbc.data", header = FALSE)
```

### 1. Default Fitting with Progress Tracking
ECM clustering is mathematically heavy. Invoking `progressr` hooks onto internal progress loops letting you monitor exactly how closely the algorithm is converging visually. Set `verbose=TRUE` to expose ECM sequential likelihood readouts.

```{r example_1, eval=FALSE}
# Enable Progress bars
handlers(global = TRUE)
handlers("cli")

# R-vine copula based mixture model over 2 classifications
with_progress({
  fit <- vcmm(data=data_wisc[,c(15,27,29,30)], total_comp=2, verbose=TRUE)
})

# Display model statistics
print(fit)
#> logLik = 353.88   BIC = -555.79   ICL = -551.16   npars = 33   initial clustering = kmeans   ECM iterations = 24

# Extract fitted dynamics (vine tree setups, parameters)
summary(fit) 
```


### 2. Evaluating Predicted Densities
Now that the arrays are structured dynamically through `rvinecopulib`, density predictions access objects strictly formatted via the list output nested inside `fit$output$vine_models`. 

```{r example_2, eval=FALSE}
# Evaluate the density of the fitted model at a specific dimensional boundary
densities <- dvcmm(c(2.747, 0.1467, 0.13, 0.05334), 
                   margin = fit$output$margin, 
                   margin_pars = fit$output$marginal_param, 
                   RVMs = fit$output$vine_models,     # Note: these are native rvinecopulib structures!
                   mix_probs = fit$output$mixture_prob)
                   
print(densities)
```

### 3. Deploying Advanced Nonparametric Copulas
To maximize detection across variables that map unexpectedly, insert `"nonparametric"` alongside parametric sets.

```{r example_3, eval=FALSE}
fit_np <- vcmm(data=data_wisc[, c(15,27,29,30)], 
               total_comp=2, 
               bicop=c("clayton", "gumbel", "nonparametric"))  
```


### 4. Direct Initial Routing
Launch different partition rules to jump-start ECM dependencies minimizing iterations correctly.

```{r example_4, eval=FALSE}
# Use hierarchical trees or GMMs as initial mapping anchors
fit_gmm <- vcmm(data=data_wisc[,c(15,27,29,30)], total_comp=2, methods=c("gmm"))  
fit_hcVVV <- vcmm(data=data_wisc[,c(15,27,29,30)], total_comp=2, methods=c("hcVVV"))  
``` 

## Contact

Please contact <O.Sahin@tudelft.nl> if you have any questions.

## References

Sahin, {"O}., & Czado, C. (2022). Vine copula mixture models and clustering for non-Gaussian data. Econometrics and Statistics. <doi:10.1016/j.ecosta.2021.08.011>.
[preprint](https://arxiv.org/pdf/2102.03257.pdf), [article](https://doi.org/10.1016/j.ecosta.2021.08.011)
